---
title: "Final Project - Stat 425"
author: "Anurag Anand"
date: "12/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(faraway)
library(ggplot2)
library(ggmap)
library(RColorBrewer)
library(tidyverse)
library(caret)
library(ggcorrplot)
library(pls)
library(randomForest)
library(rpart)
library(rpart.plot)
```

#### Reading Data

```{r c1}

# Reading the data frame
df=readxl::read_xlsx("Real_estate_valuation_data_set.xlsx")
head(df,5)
```


#### Data Preparation

```{r c2}
# Changing column names
colnames(df) = c("index",
                 "transaction_date",
                 "house_age",
                 "distance_to_nearest_metro_station",
                 "no_of_convenience_stores",
                 "latitude",
                 "longitude",
                 "house_price")

# dropping index
df = df[,-1]

# Sorting the months (The dates range from Aug 2012 to July 2013)
months = sort(unique(df$transaction_date))

# subsetting months vector
v1 = months[1:5] # year 2012
v2 = months[6:12] # year 2013

# helper function find out month number
month = function(x){
  
  if(x == months[1]){return(08)}
  else if(x == months[2]){return(09)}
  else if(x == months[3]){return(10)}
  else if(x == months[4]){return(11)}
  else if(x == months[5]){return(12)}
  else if(x == months[6]){return(01)}
  else if(x == months[7]){return(02)}
  else if(x == months[8]){return(03)}
  else if(x == months[9]){return(04)}
  else if(x == months[10]){return(05)}
  else if(x == months[11]){return(06)}
  else if(x == months[12]){return(07)}
}

# helper function to determine year
year = function(x){
  if(x %in% v1){return(2012)}
  else if(x %in% v2) {return(2013)}
}


df$month = sapply(df$transaction_date, month) # creating month column
df$year = sapply(df$transaction_date, year) # creating year column

# Dropping transaction_date column
df = df[,-1]
df = df[,c("year","month","house_age","distance_to_nearest_metro_station","no_of_convenience_stores","latitude","longitude","house_price")]
head(df,5)

```

#### Data Summary

```{r}
df_summary = df %>% mutate(year = factor(year),
                           month = factor(month))

summary(df_summary)
for(item in colnames(df)){
  str = paste0("is.na(df$",item,")")
  print(paste0("Number of Missing values in the column ",item," is = ",sum(eval(parse(text = str)))))
}

```



#### Univariate Analysis

```{r}
# Starting univariate analysis
# Distribution of house_age
ggplot(df, aes(x=house_age)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="#0073C2FF", binwidth = 1)+
  geom_density(alpha=.2, fill="#FF6666") + 
  labs(title = "Distribution of House Age", x = "House Age", y = "Proportion") + 
  geom_vline(aes(xintercept=mean(house_age)),color="blue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(house_age)),color="black", linetype="solid", size=1) +
  annotate(geom = "text",
           label = c("mean", "median"),
           x = c(mean(df$house_age), median(df$house_age)-1.5),
           y = c(0.06, 0.06),
           angle = 90, 
           vjust = 1)


# Distribution of distance_to_nearest_metro_station
ggplot(df, aes(x=distance_to_nearest_metro_station)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="#0073C2FF", binwidth = 500)+
  geom_density(alpha=.2, fill="#FF6666") + 
  labs(title = "Distribution of Distance to Metro Station", x = "Distance", y = "Proportion") + 
  geom_vline(aes(xintercept=mean(distance_to_nearest_metro_station)),color="blue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(distance_to_nearest_metro_station)),color="black", linetype="solid", size=1) + 
  annotate(geom = "text",
           label = c("mean", "median"),
           x = c(mean(df$distance_to_nearest_metro_station), median(df$distance_to_nearest_metro_station)),
           y = c(0.00075, 0.00075),
           angle = 90, 
           vjust = 1)


# Frequency plot of Number of Houses
ggplot(df %>% group_by(no_of_convenience_stores) %>% summarise(Houses = n()), aes(x = no_of_convenience_stores, y = Houses)) +
  geom_bar(fill = "#0073C2FF", stat = "identity") +
  geom_text(aes(label = Houses), vjust = -0.3) +
  labs(title = "Number of Houses vs Number of Convenience Stores")



# Distribution of price
ggplot(df, aes(x=house_price)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="#0073C2FF", binwidth = 5)+
  geom_density(alpha=.2, fill="#FF6666") + 
  labs(title = "Distribution of House Price", x = "House Price", y = "Proportion") + 
  geom_vline(aes(xintercept=mean(house_price)),color="blue", linetype="dashed", size=1) +
  geom_vline(aes(xintercept=median(house_price)),color="black", linetype="solid", size=1) +
  annotate(geom = "text",
           label = c("mean", "median"),
           x = c(mean(df$house_price), median(df$house_price)-3),
           y = c(0.04, 0.04),
           angle = 90, 
           vjust = 1)


```

#### Bivariate Analysis

```{r}
# Add the regression line
ggplot(df, aes(x=house_age, y=house_price)) + 
  geom_point()+
  geom_smooth(method=lm) +
  labs(title = paste0("House Price vs House Age (Corr = ",as.character(round(cor(df$house_age, df$house_price),2)),")"), x = "House Age", y = "House Price")


ggplot(df, aes(x=distance_to_nearest_metro_station, y=house_price)) + 
  geom_point()+
  geom_smooth(method=lm) +
  labs(title = paste0("House Price vs Distance to Nearest Metro Station (Corr = ",
                      as.character(round(cor(df$distance_to_nearest_metro_station, 
                                             df$house_price),2)),")"), 
       x = "Distance to Nearest Metro Station", 
       y = "House Price")

# House Price vs. Number of Convenience Stores
ggplot(df %>% 
         group_by(no_of_convenience_stores) %>% 
         summarise(avg_price = mean(house_price)) %>% 
         mutate(avg_price = round(avg_price,2)), aes(x = no_of_convenience_stores, y = avg_price)) +
  geom_bar(fill = "#0073C2FF", stat = "identity") +
  geom_text(aes(label = avg_price), vjust = -0.3) +
  labs(title = "Average Price by different number of nearby convenience stores", x = "No of Stores", y = "Average House Price")



# Making a scatter plot of lon
ggplot(df, aes(x=longitude, y=latitude, colour=house_price)) +
  geom_point(alpha = 0.3, size = 3.5) + 
  scale_colour_gradient(low="#A9C8F3", high="#0C2389")+
  labs(title = "House Price vs Location")


# Checking variation of price against months

# Boxplot basic

box_df = df %>% 
  select(year, month, house_price) %>% 
  arrange(year, month) %>% 
  mutate(year_month = factor(as.Date(paste0(year,"-",month,"-01")))) %>% 
  select(year_month, house_price)

ggplot(box_df, aes(x=year_month, y=house_price)) + 
  geom_boxplot(fill="slateblue", alpha=0.2) + 
  labs(title = "Variation of House Price by Year Month", x = "Year-Month", y = "House Price") + 
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1))



# Correlation Matrix
ggcorrplot(round(cor(df[,colnames(df)[3:8]]),2), hc.order = TRUE, type = "full",
           outline.col = "white",
           ggtheme = ggplot2::theme_gray,
           colors = c("#E46726", "white", "#6D9EC1"))

```

#### Leverage Points and Outlier

```{r}
# Identifying high leverage points
lmod = lm(house_price ~ ., data = df)
lev = influence(lmod)$hat
sort(lev[lev > 3*ncol(df)/nrow(df)])

halfnorm(lev, 8, labs = row.names(df), ylab = "leverages")


max(cooks.distance(lmod))
halfnorm(cooks.distance(lmod), labs = row.names(df), ylab = "Cook's distance")

# deleting rows
df = df[-271,]
df = df[-313,]
```

Deleting row number 217 and 313

#### Modelling

Linear Regression with Prediction on Test Set

```{r}
df_clean = df

# Test train split

df_trn_idx = sample(nrow(df_clean),size = 0.8*nrow(df_clean))
df_trn = df_clean[df_trn_idx,]
df_tst = df_clean[-df_trn_idx,-8]
y_tst = df_clean[-df_trn_idx,8]

lm1 = lm(house_price ~., data = df_trn)
summary(lm1)

# conducting prediction on the test set
calc_rmse = function(y,pred){
  sqrt(mean((y - pred)^2))
}

# Error on test data set simple linear regression
lin_mod_rmse = calc_rmse(as.vector(y_tst$house_price),predict(lm1, df_tst))
print("Linear Regression RMSE:")
print(lin_mod_rmse)

# Regression Diagnostics
plot(lm1,1)
plot(lm1,2)
plot(lm1,3)
plot(lm1,4)

```



Principal Component Regression

```{r}
pr_house = prcomp(df_trn[,1:7], center = TRUE, scale = TRUE)
summary(pr_house)
dim(pr_house$x)


scree_df = data.frame(cbind(1:7,pr_house$sdev))
colnames(scree_df) = c("PC Number","SD of PC")
ggplot(data=scree_df, aes(x=`PC Number`, y=`SD of PC`, group=1)) +
  geom_line()+
  geom_point() + 
  labs(title = "Scree Plot for PCR")


# Lets find out the number of components required using cross validation
pcrmod = pcr(house_price ~. , data = df_trn, ncomp = 7)
pcrmse = RMSEP(pcrmod, newdata = cbind(df_tst,y_tst))

cv_df = data.frame(cbind(0:7,pcrmse$val))
colnames(cv_df) = c("Number of Components","RMSE")
ggplot(data=cv_df, aes(x=`Number of Components`, y=`RMSE`, group=1)) +
  geom_line()+
  geom_point() + 
  labs(title = "Test Error vs Number of Components")

# Finding out what number of Principal Components give the least error on the test set 
which.min(pcrmse$val)

# Training model with 5 principal components
pr_house = prcomp(df_trn[,1:7], center = TRUE, scale = TRUE)
summary(pr_house)
pcr_train_df = cbind(data.frame(pr_house$x[,1:5]),df_trn$house_price)
colnames(pcr_train_df) = c("PC1","PC2","PC3","PC4","PC5","house_price")
pcr_mod = lm(house_price ~ ., data = pcr_train_df)
summary(pcr_mod)
```
Summary Statistics:
  - 96% of variation captured with 5 principle components
Scree Plot:
  - Elbow at PC4
Testing Error:
  - Lowest prediction error around 5 PC's

```{r}
# Changing the test data set
pcr_test = data.frame(predict(pr_house, df_tst)[,1:5])
pcr_rmse = calc_rmse(as.vector(y_tst$house_price),predict(pcr_mod, pcr_test))

```

#### Lasso Regression


```{r}

library(glmnet)
set.seed(42)
#perform k-fold cross-validation to find optimal lambda value
x = data.matrix(df_trn[,1:7])
y = data.matrix(df_trn[,8])
cv_model = cv.glmnet(x, y, alpha = 1)

print(cv_model$lambda.min)

plot(cv_model)

#find optimal lambda value that minimizes test MSE
best_lambda = cv_model$lambda.min


# Fitting the best model
best_model = glmnet(x, y, alpha = 1, lambda = best_lambda)

coef(best_model)

lasso_rmse = calc_rmse(as.vector(y_tst$house_price),predict(best_model, data.matrix(df_tst)))
lasso_rmse

```


#### Ridge Regression



```{r}

library(glmnet)
set.seed(42)
#perform k-fold cross-validation to find optimal lambda value
x = data.matrix(df_trn[,1:7])
y = data.matrix(df_trn[,8])
cv_model_ridge = cv.glmnet(x, y, alpha = 0)


plot(cv_model_ridge)

#find optimal lambda value that minimizes test MSE
best_lambda = cv_model_ridge$lambda.min
print(best_lambda)


# Fitting the best model
best_model_ridge = glmnet(x, y, alpha = 0, lambda = best_lambda)

coef(best_model_ridge)

# Prediction
ridge_rmse = calc_rmse(as.vector(y_tst$house_price),predict(best_model_ridge, data.matrix(df_tst)))
ridge_rmse

```

#### Decision Tree

```{r}
df_trn$year = factor(df_trn$year)
df_trn$month = factor(df_trn$month)

df_tst$year = factor(df_tst$year)
df_tst$month = factor(df_tst$month)

tree_mod = rpart(house_price ~. , data = df_trn)
rpart.plot(tree_mod)

tree_mod_rmse = calc_rmse(as.vector(y_tst$house_price),predict(tree_mod, df_tst))
tree_mod_rmse


```



#### Random Forest

```{r}
library(randomForest)
df_trn$year = factor(df_trn$year)
df_trn$month = factor(df_trn$month)

df_tst$year = factor(df_tst$year)
df_tst$month = factor(df_tst$month)

# Training model
rf_mod = randomForest(house_price ~., data = df_trn, ntree = 500, mtry = 3)

# PLotting RMSE against Number of trees
plot(rf_mod)

# Calculating test RMSE
rf_mod_rmse = calc_rmse(as.vector(y_tst$house_price),predict(rf_mod, df_tst))
rf_mod_rmse




# Get importance values as a data frame
imp = as.data.frame(importance(rf_mod))
imp = cbind(vars=rownames(imp), imp)
imp = imp %>% arrange(imp$IncNodePurity)




# Frequency plot of Number of Houses
ggplot(data = imp, aes(x = vars, y = IncNodePurity)) +
  geom_bar(fill = "#0073C2FF", stat = "identity") +
  geom_text(aes(label = round(IncNodePurity,2)), vjust = -0.3) +
  labs(title = "Variable Importance Plot - Random Forest") + 
  theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1))
```

```{r}
# Model Accuracies
acc_df = data.frame(cbind(c("Linear Regression","PCR Regression","Lasso Regression","Ridge Regression","Decision Tree", "Random Forest"),
               c(lin_mod_rmse, pcr_rmse, lasso_rmse, ridge_rmse, tree_mod_rmse, rf_mod_rmse)))


colnames(acc_df) = c("Model Name", "RMSE")
acc_df$RMSE = round(as.numeric(acc_df$RMSE),2)

ggplot(data = acc_df, aes(x = `Model Name`, y = `RMSE`)) +
  geom_bar(fill = "#0073C2FF", stat = "identity") +
  geom_text(aes(label = `RMSE`), vjust = -0.3) +
  labs(title = "Test RMSE for Different Models") + 
  theme(axis.text.x = element_text(angle=90, vjust=1, hjust=1))
```

